Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-06 paper.tex
Update DTWF section

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\SLiM}[0]{\texttt{SLiM}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
\newcommand{\phastsim}[0]{\texttt{phastsim}}
\newcommand{\discoal}[0]{\texttt{discoal}}

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}

@@

@@

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
has been integral to theoretical developments since its beginnings.
Because of the ease with which the basic coalescent model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. Moreover, \msprime\ is a community oriented open-source development
project, aimed as a way to reduce duplication of effort and increase
the quality of key simulation software.
\end{abstract}

@@

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0, a state-of-the-art
coalescent simulator both with respect to efficiency as well as feature-richness.
Here we provide an overview of the extensive abilities of \msprime\ 1.0,
demonstrate through numerous performance comparisons its advantages over
current, alternative software packages, and finally lay out future plans for
further open-source development of what we hope will become an industry
standard piece of computational infrastructure for the genetics community.

@@

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into
a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

@@

The tree sequence encoding of ancestry as set of nodes and edges
as just described, is equivalent to a graph in which edges
are annotated with a set of disjoint genomic intervals.
This is closely related to the Ancestral Recombination Graph (ARG),
a term that is often loosely understood
to mean some representation of genetic
ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
In this sense a succinct tree sequence \emph{is} an ARG.
More technically, an ARG is a particular
representation of genetic ancestry as a graph based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although
the default minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section).
It should be noted that, despite standardisation
efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

@@

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it very easy to run simple
simulations, but as model complexity and the number of parameters specified increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from some distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and
a number of packages have been developed
to allow simulations be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

@@

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows. The
\texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulating complex demography models straightforward.

@@

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
Numerous forks  of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

@@

\begin{figure}
    \includegraphics{illustrations/unmutated_tree}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualisation of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
    }
\end{figure}

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. Although \msprime\ also supports simulating both ancestry and mutations,
the two aspects are functionally independent within the software.
Both ancestry and mutation simulators are present in \msprime
for reasons of convenience and history,
and could be split into separate packages.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

As well as providing a new API (see the \nameref{sec-sim-interface} section)
which emphasises the logical split between ancestry and mutation simulation,
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We  support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
so that other models, such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation},
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

\begin{figure}
    \includegraphics[width=\textwidth]{figures/mutations-perf}
\caption{\label{fig-mutations-perf} Time required to run
\texttt{sim\_mutations} on tree sequences generated
by \texttt{sim\_ancestry} (with a population size of $10^4$
and recombination rate of $10^{-8}$) for varying (haploid) sample
size and sequence length. We run 10 replicate mutation simulations
each for three different mutation rates, and report the average
CPU time required (Intel Core i7-9700).
(A) Holding sequence length fixed at 10 megabases and varying the
number of samples (tree tips) from 10 to 100,000.
(B) Holding number of samples fixed at 1000, and varying the sequence
length from 1 to 100 megabases.}
\end{figure}

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis
(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
appendix),
rather than tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
(since there are 137,081 trees and 1,998 branches in each
tree) rather than 508,125 edges, resulting in $\sim500$ times more work.
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(negating the advantage of working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods, because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
Given these timings, it seems quite unlikely
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

There are many ways in which the mutation generation code
in \msprime\ could be extended. For example, we intend to add support for
microsatellites~\citep{mailund2005coasim},
codon models~\citep{arenas2007recodon}
and indels~\citep{cartwright2005dna,fletcher2009indelible},
although some may also require upstream additions to \tskit's data model
which is currently based on the assumption of independent sites.

@@

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time, generating common ancestor and recombination
events and tracking their effects on segments of ancestral
material inherited from the
sample~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events merge the ancestral material of two lineages,
and result in coalescences in the marginal trees when ancestral
segments overlap.
Recombination events split the ancestral material for some lineage
at a given breakpoint, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots and patterns
of recombination from empirical maps.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and simulations by \ms, \texttt{msHot} and \SLiM.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Running time for \msprime\
    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
    Each point is the run time of one simulation,
    for various values of effective population size ($N_e$),
    genome length in Morgans ($L$), and number of samples ($n$).
    Run time scales quadratically with the product of $N_e$ and $L$,
    shown on the horizontal axis. For example, (A) shows that
    1,000 samples of 1 Morgan-length chromosomes
    from a population of $N_e=2,000$ diploids
    would take about 2 seconds, and (equivalently) that the same number of
    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
    Since recombination rate in these simulations was $10^{-8}$,
    $L$ is the number of base pairs divided by $10^8$.
    The black lines are quadratic fits separately in each panel
    and sample size. Also shown are the approximate values of
    $N_e L$ for chromosome 1 in three species, using values from
    the \stdpopsim\ catalogue~\citep{adrion2020community}.
}
\end{figure}

@@

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
At the maximum, coalescence and recombination rates are equal,
so a typical segment of ancestry will spend roughly half its time
in a lineage with at least one other such segment---and,
since such lineages carry at least two segments,
at most one-third of the lineages carry long trapped segments of ancesry.
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
this implies that $\ell \approx L / 6$.
Setting the rates of recombination and common ancestor events
to be equal and solving for $M$,
we find that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 6$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

@@

\subsection*{Gene conversion}

@@

\end{center}
\caption{\label{fig-arg} (A) An simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation;
(D) the fraction of nodes in a tree sequence that are concerned with
extra ARG information in a tree sequences returned using \msprime's
\texttt{record\_full\_arg} mode}
\end{figure}

>>>>

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\msHOT}[0]{\texttt{msHOT}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\SLiM}[0]{\texttt{SLiM}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
\newcommand{\phastsim}[0]{\texttt{phastsim}}
\newcommand{\discoal}[0]{\texttt{discoal}}
\newcommand{\ARGON}[0]{\texttt{ARGON}}

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
\newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}

@@

% \author[n]{Graham Gower}
% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}

@@

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
has been integral to theoretical developments since its beginnings.
Because of the ease with which the basic coalescent model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. Moreover, development of \msprime\ is collaborative
and open source, which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

@@

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0, a state-of-the-art
coalescent simulator both with respect to efficiency as well as feature-richness.
Here we provide an overview of the extensive abilities of \msprime\ 1.0,
demonstrate its performance advantages over alternative software,
and finally lay out future plans for
further open-source development of what we hope will become an industry
standard piece of computational infrastructure for the genetics community.

@@

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into
a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

@@

The tree sequence encoding of ancestry as a set of nodes and edges
as just described, is equivalent to a graph in which edges
are annotated with a set of disjoint genomic intervals.
This is closely related to the Ancestral Recombination Graph (ARG),
a term that is often loosely understood
to mean some representation of genetic
ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
In this sense a succinct tree sequence \emph{is} an ARG.
More technically, an ARG is a particular
representation of genetic ancestry as a graph based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although
the default minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section).
It should be noted that, despite standardisation
efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

@@

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it very easy to run simple
simulations, but as model complexity and the number of parameters specified increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from some distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and
a number of packages have been developed
to allow simulations to be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

@@

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows. The
\texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.

@@

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
Numerous forks  of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \msHOT~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

@@

\begin{figure}
    \includegraphics{illustrations/unmutated_tree}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualisation of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
        \grgcomment{Add \texttt{sim\_ancestry} and \texttt{sim\_mutations} as subfigure titles?}
    }
\end{figure}

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. Although \msprime\ also supports simulating both ancestry and mutations,
the two aspects are functionally independent within the software.
Both ancestry and mutation simulators are present in \msprime\
for reasons of convenience and history,
and could be split into separate packages.
The mature and efficient C and Python interfaces for
\tskit\ make it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

As well as providing a new API
that emphasises the logical split between ancestry and mutation simulation
(see the \nameref{sec-sim-interface} section),
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
Other models, such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation},
can be defined easily and efficiently in
user code by specifying an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings).
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
\Pyvolve~\citep{spielman2015pyvolve}.

\begin{figure}
    \includegraphics[width=\textwidth]{figures/mutations-perf}
\caption{\label{fig-mutations-perf} Time required to run
\texttt{sim\_mutations} on tree sequences generated
by \texttt{sim\_ancestry} (with a population size of $10^4$
and recombination rate of $10^{-8}$) for varying (haploid) sample
size and sequence length. We ran 10 replicate mutation simulations
each for three different mutation rates, and report the average
CPU time required (Intel Core i7-9700).
(A) Holding sequence length fixed at 10 megabases and varying the
number of samples (tree tips) from 10 to 100,000.
(B) Holding number of samples fixed at 1000, and varying the sequence
length from 1 to 100 megabases.}
\end{figure}

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis
(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
appendix),
rather than tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
(since there are 137,081 trees and 1,998 branches in each
tree) rather than 508,125 edges, resulting in $\sim$500 times more work.
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(negating the advantage of working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods, because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
Given these timings, it seems quite unlikely
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

There are many ways in which the mutation generation code
in \msprime\ could be extended. For example, we intend to add support for
microsatellites~\citep{mailund2005coasim},
codon models~\citep{arenas2007recodon}
and indels~\citep{cartwright2005dna,fletcher2009indelible},
although changes may be required to \tskit's data model
which is currently based on the assumption of independent sites.

@@

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time, generating common ancestor and recombination
events and tracking their effects on segments of ancestral
material inherited from the
sample~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events merge the ancestral material of two lineages,
and result in coalescences in the marginal trees when ancestral
segments overlap.
Recombination events split the ancestral material for some lineage
at a given breakpoint, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots and patterns
of recombination from empirical maps.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and simulations by \ms, \msHOT\ and \SLiM.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Running time for \msprime\
    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
    Each point is the run time of one simulation,
    for various values of effective population size ($N_e$),
    genome length in Morgans ($L$), and number of samples ($n$).
    Run time scales quadratically with the product of $N_e$ and $L$,
    shown on the horizontal axis. For example, (A) shows that
    1,000 samples of 1 Morgan-length chromosomes
    from a population of $N_e=2,000$ diploids
    would take about 2 seconds, and (equivalently) that the same number of
    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
    Since recombination rate in these simulations was $10^{-8}$,
    $L$ is the number of base pairs divided by $10^8$.
    The black lines are quadratic fits separately in each panel
    and sample size. Vertical grey lines show the approximate values of
    $N_e L$ for chromosome 1 in three species, using values from
    the \stdpopsim\ catalogue~\citep{adrion2020community}.
}
\end{figure}

@@

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
At the maximum, coalescence and recombination rates are equal,
so a typical segment of ancestry will spend roughly half its time
in a lineage with at least one other such segment---and,
since such lineages carry at least two segments,
at most one-third of the lineages carry long trapped segments of ancestry.
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
this implies that $\ell \approx L / 6$.
Setting the rates of recombination and common ancestor events
to be equal and solving for $M$,
we find that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 6$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

@@

\subsection*{Gene conversion}
\grgcomment{TODO: discuss Fig~\ref{fig-gc-perf} in text of GC section.}

@@

\end{center}
\caption{\label{fig-arg} (A) A simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation;
(D) the fraction of nodes in a tree sequence that are concerned with
extra ARG information in a tree sequence returned using \msprime's
\texttt{record\_full\_arg} mode}
\end{figure}
